---
title: "SMDE: Second Assignment"
author: "Adrià Lisa"
date: "2023-12-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## 1. Defining our own RNG
We define a function to define $n$ pseudo-random numbers between $[0,1)$ with a given *seed*, for reproducibility.
A congruential generator was used, with the following hyperparameters:

* Module $m = 2^32$.
* Multiplicative factor $a = 4 \cdot 777 + 1$.
* Additive factor $c = 50033$.

A list of RN is generated via the recurrence $Z_i = a Z_{i-1} + c \,\, (mod \,m)$, where $Z_0$ is the value of the *seed*, set by default to the prime number $26107$.
```{r}
My_RNG <- function (n, seed=26107) {
  #Congruent mixt sequence of numbers (best if c and Z0 are prime)
  m = 2^32; a = 4*777+1; c = 50033 ;
  R = integer(n); R[1] <- (a*seed + c)%%m 
  for (i in 1:(n-1)){
    R[i+1] <- (a*R[i] + c)%%m
  }
  R = R / m # in [0, 1)
  return(R)
}
```
Our teacher (Pau Fonseca) explained in class that although it is not mathematically proven, it has been empirically observed that choosing the seed to be a prime number often leads to better results.

As part of our validation process, we tested our RNG with different empirical tests from the ```{r} randtests``` library, obtaining a reasonably high confidence in the hypothesis of "randomness".

```{r}
R = My_RNG(10000)
library(randtests)
bartels.rank.test(R)
turning.point.test(R)
cox.stuart.test(R)
difference.sign.test(R)
rank.test(R)
```
## 2. Simulating our data
We created a table of numeric data, using R's standard RNG, and save it in the file "rngdata.csv"

```{r}
# Set seed for reproducibility
set.seed(123)

# Number of individuals and factors
num_individuals <- 2000
num_factors <- 10

# Create an empty data frame
data <- data.frame(matrix(NA, nrow = num_individuals, ncol = num_factors + 1))
colnames(data) <- c(paste("Factor", 1:num_factors, sep = ""), "Answer")

# Define distributions for Factor 1 to 5
for (i in 1:5) {
  data[, i] <- runif(num_individuals, min=0, max = 10)
}

# Define linear functions for Factor 6 to 10

data$Factor6 <- 6*data$Factor1 + data$Factor3
data$Factor7 <- data$Factor1 + data$Factor2 + data$Factor3 + data$Factor4 + data$Factor5
data$Factor8 <- data$Factor1 - data$Factor5
data$Factor9 <- 3*data$Factor3 - 2*data$Factor4
data$Factor10 <- -7*data$Factor2 - 5*data$Factor4

# Answer variable is just the sum of all factors + noise
data$Answer <- rowSums(data[,4:6]) + rnorm(num_individuals)
write.csv(data, "rngdata.csv", row.names = FALSE)


```

## 3. Obtain an expression to generate new data
Everithing here is just an output of chatgpt, it has not been reviewed and thus it can be considered to be all wrong.
```{r}
# Load the required libraries
library(tidyverse)

# Read the CSV file
data <- read.csv("rngdata.csv")

# Correlation matrix
cor_matrix <- cor(data)
print("Correlation Matrix:")
print(cor_matrix)

# Simple linear regression for each factor vs. Answer
for (i in 1:10) {
  lm_model <- lm(Answer ~ data[, i], data = data)
  print(paste("Linear Regression for Factor", i, "vs. Answer:"))
  print(summary(lm_model))
}
```

We now see what happens when we build a multiple linear regression (MLR), with all the factors
```{r}
# Multiple linear regression
mlr_model <- lm(Answer ~ ., data = data)
print("Multiple Linear Regression:")
print(summary(mlr_model))
```
Since only the factors from 1 to 5 are independent variables, they are the only ones that are found relevant for the MLR. 


```{r}
# Plot residuals vs. fitted values for the multiple linear regression
plot(mlr_model$fitted.values, mlr_model$residuals, main = "Residuals vs. Fitted", xlab = "Fitted Values", ylab = "Residuals")


# Anova for the multiple linear regression
anova_result <- anova(mlr_model)
print("ANOVA for Multiple Linear Regression:")
print(anova_result)

# Plot the residuals to check for normality
hist(mlr_model$residuals, main = "Histogram of Residuals", xlab = "Residuals")

# Q-Q plot for the residuals
qqnorm(mlr_model$residuals)
qqline(mlr_model$residuals)

# Boxplot of residuals
boxplot(mlr_model$residuals, main = "Boxplot of Residuals")

# Check for homoscedasticity
plot(mlr_model$fitted.values, abs(mlr_model$residuals), main = "Residuals vs. Fitted Values (Homoscedasticity Check)", xlab = "Fitted Values", ylab = "Absolute Residuals")

# Additional exploratory analysis as needed

# Save plots or results as needed
# Example: save the scatterplot matrix as an image file
png("scatterplot_matrix.png")
pairs(data)
dev.off()

```

However, we believe that we can build a lighter model, that contains less than 5 components. To do so, we now perform a principal component analysis (PCA).

```{r}
pca<-prcomp(data[, -ncol(data)])
summary(pca)
```

Since we can achive 88% explainablity by just taking the first 2 components, we will build a linear model from PC1 and PC2. 
```{r}
pca_data <- as.data.frame(cbind(pca$x[,1:2], data$Answer))
pca_model <- lm(V3 ~ ., data = pca_data)
summary(pca_model)

```
All the test statistics of this models are also very good, so we will prefer this one over the MLR with all the factors because it is lighter. 

Even so, we will now make some tests on this model

1. First, the error term must be normally distributed:
```{r}
qqnorm(residuals(pca_model))
```
The qq-plot shows that this is indeed the case.

2. The errors should also have homogeneous variance. The Breusch-Pagan test gives a p-value greater than $0.05$, so we accept the hypothesis of homoscedasticity.
```{r}
library(lmtest)
bptest(pca_model)
```

3. The errors should also be independent. The Durbin-Watson statistic is very close to $2.0$, thus, we can say there is no autocorrelation in the residuals.
```{r}
dwtest(pca_model)
```

We need not test if there is correlation among the four selected principal components, because this holds by definition.


```{r}

queue_time1 <- pca_model$coefficients[2]*pca_data$PC1
mean(queue_time1)
min(queue_time1)
sd(queue_time1)
pca_model$coefficients[3]
queue_time2 <- pca_model$coefficients[3]*pca_data$PC2
mean(queue_time2)
min(queue_time2)
sd(queue_time2)
```

```{r}
hist(data$Answer)
```
```{r}
###Operational Validation###
n <- nrow(pca_data)
data.sample1 <- sample(1:n, round(0.67*n))
data.set1 <- pca_data[data.sample1, ] 
test.set1 <- pca_data[-data.sample1, ] 
data.model1 <- pca_model

?predict
yhat<-predict(data.model1, data.set1, interval="prediction")
yhat

y<-test.set1$Answer

error<-cbind(yhat[,1,drop=FALSE],y,(y-yhat[,1])^2)
sqr_err<-error[,1]
mse<-mean(sqr_err)

### Root Mean Square Error ###
RMSE1<-sqrt(mse/(nrow(test.set1)))
RMSE1

names(data.model1)
RMSE_data1<- sqrt(mean((pca_model$residuals)^2)/nrow(data.set1))
RMSE_data1


```

3.Define a DOE(Definition of Experiments) to explore with the parametrization of the 10 factors the answer obtains the best value (we consider the minimum one to be the best value).

```{r}
exp <- data$Answer-1
model <- aov(data$Answer~exp)
summary.aov(model)

```

```{r}
eval_with_pca <- function(factors, rotation, pca_model) {
  pca_factors <- rotation %*% unlist(factors)
  answer = pca_model$coefficients %*% c(1,pca_factors[c(1,2)]) 
  return(answer)
}

neutral <- function(data){
  sdevs <- apply(data, 2, sd)
  means <- apply(data ,2, mean)
  rand <-rnorm(10,means, sdevs)
  return(rand)
}

run_experiments <- function(){
  
  data_min <- apply(data, 2, min)
  data_max <- apply(data, 2, max)
  doe <-(data[1,])
  doe[1,1:10] <- neutral(data)
  doe <- cbind(Level="neutral", doe)
  
  doe[1, ]$Answer <- eval_with_pca(doe[1,-c(1,12)], pca$rotation, pca_model)
  
  for (i in 1:10) {
    newdata<-neutral(data)
    names(newdata)<-NULL
    newdata[i]<-data_min[i]
    ans <- eval_with_pca(newdata, pca$rotation, pca_model)
    row<-c(paste("-", i), newdata, ans)
    doe[2*i,]<-row
    newdata[i]<-data_max[i]
    ans <- eval_with_pca(newdata, pca$rotation, pca_model)
    row<-c(paste("+", i), newdata, ans)
    doe[2*i+1,]<-row
  
  }
  for(j in 1:9){
    for( k in (j+1):10){
    newdata<-neutral(data)
    newdata[j]<-data_min[j]
    newdata[k]<-data_min[k]
    ans <- eval_with_pca(newdata, pca$rotation, pca_model)
    row<-c(paste("-_-", j,k), newdata, ans)
    doe <- rbind(doe,row)
    newdata[j]<-data_min[j]
    newdata[k]<-data_max[k]
    ans <- eval_with_pca(newdata, pca$rotation, pca_model)
    row<-c(paste("-_+", j,k), newdata, ans)
    doe <- rbind(doe,row)
    newdata[j]<-data_max[j]
    newdata[k]<-data_min[k]
    ans <- eval_with_pca(newdata, pca$rotation, pca_model)
    row<-c(paste("+_-", j,k), newdata, ans)
    doe <- rbind(doe,row)
    newdata[j]<-data_max[j]
    newdata[k]<-data_max[k]
    ans <- eval_with_pca(newdata, pca$rotation, pca_model)
    row<-c(paste("+_+", j,k), newdata, ans)
    doe <- rbind(doe,row)
    }
  }
  return(doe[c(1,12)])
}

```

ANOVA tests for 1-factor and 2-factor

```{r}

subset_data <- doe[2:201, ]
# 1-factor ANOVA
for (i in 1:10) {
  subset_data[, i] <- as.factor(subset_data[, i])
  doe_model <- aov(Answer ~ subset_data[, i+1], data = subset_data)
  print(paste("One-Factor ANOVA for Factor", i, "vs. Answer:"))
  print(summary(doe_model))
}

```
```{r}
subset_data2 <- doe[2:144, ] 

for (i in 1:9) {
  for (j in (i + 1):10) {
    subset_data2[, 1:10] <- lapply(subset_data2[, 1:10], as.factor)
    # Two-Factor ANOVA
    doe_model <- aov(Answer ~ subset_data2[, i+1] + subset_data2[, j+1], data = subset_data2)
    # Print ANOVA summary
    print(paste("Two-Factor ANOVA for Factor", i, "and Factor", j, "vs. Answer:"))
    print(summary(doe_model))
  }
}

```